Parsing Postfix log files

### 2009-03-25 doc/logparser-implementation.tex
Corrections from proofreading, with lots more to come.

@@

Please refer to \sectionref{parser design} for a discussion of why the
rules, actions, and framework have been separated in the parser's design.
In this section algorithm can be taken to mean the combination of framework
and actions.

@@

    \item Postfix logs enough information to make it possible to accurately
        reconstruct the actions it has taken.  There are several heuristics
        used when parsing; see \sectionref{identifying bounce
        notifications}, \sectionref{aborted delivery attempts}, and
        \sectionref{pickup logging after cleanup} for details.

@@

In some ways this task is similar to reverse engineering or replicating a
black box program based solely on its inputs and outputs.  Thus far
analysis of the log files has been sufficient to reconstruct Postfix's
behaviour, but for more difficult programs the techniques described in
\cite{black-box-error-reporting} may be useful.  There are some advantages
to treating Postfix as a black box during parser development:

@@

    \item The parser is developed using real world log files rather than
        the idealised log files someone would naturally envisage reading
        the source code.  The source code cannot accurately communicate the
        variety of orderings in which log lines are written to the log
        file, as process scheduling independently interferes with logging
        and other processing.

    \item The parser acts as a second source of information about Postfix's
        operation, using information gathered from empirical evidence.  A
        separate project could compare the empirical knowledge inherent in
        the parsing algorithm with Postfix's documentation and source code
        to see how closely the two agree.

@@

A picture is said to be worth a thousand words: the flow chart in
\figureref{flow chart image} shows the most common paths a connection or
mail can take through \parsername{}; the parsing complications described in
\sectionref{complications} are excluded for the sake of clarity.  The flow
chart is intended to be a graphical overview of how a mail progresses
through both Postfix and \parsername{}, providing an overall context into
which the detailed descriptions in the remainder of this chapter will fit,
in particular the actions (\sectionref{actions in implementation}) and
complications (\sectionref{complications}).

@@

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop}, by \acronym{SMTP}, by re-injection due to
forwarding, or internally generated by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately and the sender address is
logged \flowchart{PICKUP}{2}.  Re-injection due to forwarding sadly lacks
explicit log lines of its own; it is explained fully in
\sectionref{Re-injected mails}.  Internally generated mails lack any
explicit origin in Postfix 2.2.x and must be detected using heuristics as
described in \sectionref{identifying bounce notifications}; later versions
of Postfix do provide log lines for internally generated mails.  Bounce
notifications are the primary example of internally generated mails, though
there are other types, e.g.\ Postfix may generate mails to the
administrator when it encounters configuration errors, but such mails are
presumably rare.

@@

    \item The client disconnects \flowchart{DISCONNECT}{6}.  If Postfix has
        rejected any \acronym{SMTP} commands the data will be saved to the
        database; if not there will not be any data to save (any mails
        accepted will already have been cloned so their data is in another
        data structure).

    \item If one or more mails were accepted there will be more log lines
        for those mails.

@@

    \item Postfix will log the sender and recipient addresses separately
        \flowchart{SAVE\_DATA}{9}.

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail \flowchart{TRACK}{10}; the handling of
        re-injected mails is described in \sectionref{tracking re-injected
        mail}.

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator \flowchart{COMMIT}{12}.  This is the last log line
        for this particular mail (though it may be indirectly referred to
        if it was re-injected).  If it is neither parent nor child of
        re-injection the data is cleaned up and entered in the database,
        then deleted from the state tables.

@@

The database is an integral part of the parser presented here: it stores
the rules and the data gleaned by using those rules to parse Postfix log
files.  Understanding the database schema is important in understanding the
actions of the parser, and essential to developing further applications
which utilise the data gathered; \sectionref{database as API} describes
how the database schema functions as an \acronym{API}.

@@

The data saved from parsing the log files is also divided into two tables:
connections and results.  The connections table contains an entry for every
mail accepted and every connection where there was a rejection; the
individual fields will be described in \sectionref{connections table}.
There will be at least one entry in the results table for each entry in the
connections table; the result table's fields will be covered in detail in
\sectionref{results table}.  A diagram of the database schema is provided
in \figureref{Diagram of the Database Schema picture}, alongside the
in-depth descriptions of each table.

@@

The database populated by \parsername{} provides a simple interface to
Postfix log files.  Although the interface is a database schema rather than
a set of functions in a library, it has the same benefits as any other
\acronym{API}: it provides a stable interface, allowing code on either side
of the interface to be changed without adverse effects, as long as the
interface is adhered to.  Programs that use the database can range from the
simple examples in \sectionref{motivation} to far more complex data mining
tools.

@@

\begin{verbatim}
CREATE TRIGGER ON INSERT INTO results
    WHERE sender = 'boss@example.com'
        AND action = 'DELIVERY_REJECTED'
    SEND PANIC EMAIL TO 'postmaster@example.com';
\end{verbatim}

            \item Other tables can be added to the database, to cache
                historical, summary, or computed data.

@@

    \item \acronym{SQL} is reasonably standard and many people will already
        be familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn, e.g.\
        \urlLastChecked{http://philip.greenspun.com/sql/}{2009/02/23}.
        Although every vendor implements a different dialect of
        \acronym{SQL}, the basics are the same everywhere, analogous to the
        overall similarities and minor differences amongst Irish English,
        British English, American English, and Australian English.
        Depending on the database in use there may be tools available that
        reduce or remove the requirement to know \acronym{SQL}; several are
        available for \gls{SQLite}, the default database used by
        \parsername{}.

@@

Rules are discussed in detail in \sectionref{rules in implementation}, but
the structure of the rules table is documented here along with the other
tables in the database.  Rules are created by the user, not the parser, and
will not be modified by the parser (except for the hits and hits\_total
fields).  Rules classify the individual log lines, capturing data to be
saved in the connections and results tables, and specifying the action to
take for that log line.

The following attributes are defined for each rule:

@@

    \item [description] Something must have occurred to cause Postfix to
        log each line (e.g.\ a remote client connecting causes a connection
        line to be logged).  This field describes the event causing the log
        lines this rule matches.

    \item [restriction\_name] The restriction that caused the mail to be
        rejected.  Only applicable to rules that recognise rejection log
        lines, i.e.\ rules that have an action of
        \texttt{DELIVERY\_REJECTED}, other rules should have an empty
        string.

@@

    \item [regex] The regex to recognise log lines with, documented in
        \sectionref{regex components}.

    \item [connection\_data] Sometimes rules need to save data that is not
        present in the log line: e.g.\ setting \texttt{client\_ip} when a
        mail is being delivered to another server.  The format is:
        \newline{} \tab{} \texttt{ client\_hostname = localhost,}
        \newline{} \tab{} \tab{} \texttt{client\_ip = 127.0.0.1} \newline{}
        i.e.\ semi-colon or comma separated assignment statements, with the
        variable name on the left and the value on the right hand side.
        Any field in the connections table can be set in this way.  Commas
        and semi-colons cannot be escaped and thus cannot be included in
        data.  This feature is intended for use with small amounts of data,
        so dealing with escape sequences was deemed unnecessary.

    \item [result\_data] The result table equivalent of
        \texttt{connection\_data}.

@@

    \item [hits] This counter is maintained for every rule and incremented
        each time the rule successfully recognises a log line.  At the
        start of each run the parser sorts the rules by how frequently each
        rule recognised log lines, and at the end of the run it updates
        every rule's hits field in the database.  Assuming that the
        distribution of log lines is reasonably consistent across log
        files, ordering rules by their recognition frequency will reduce
        the parser's execution time.  Rule ordering for efficiency is
        discussed in \sectionref{rule ordering for efficiency}.

    \item [hits\_total] The total number of hits for this rule over all
        runs of the parser.

    \item [priority] This is the user-configurable companion to hits: when
        the list of rules is sorted, priority overrides hits.  This allows
        more specific rules to take precedence over more general rules
        (described in \sectionref{rules in architecture}).

@@

Every accepted mail and every connection where there was a rejection will
have a single entry in the connections table containing the following
fields:

@@

    \item [server\_ip] The \acronym{IP} address (IPv4 or IPv6) of the
        server: the local address when receiving mail, the remote address
        when sending mail.

@@

    \item [client\_ip] The client \acronym{IP} address (IPv4 or IPv6): the
        remote address when receiving mail, the local address when sending
        mail.

@@

    \item [helo] The hostname used in the HELO command.  The HELO hostname
        occasionally changes during a connection, presumably because spam
        or virus senders think it is a good idea.  By default Postfix only
        logs the HELO hostname when it rejects an \acronym{SMTP} command,
        but it is quite simple to rectify this, as described in
        \sectionref{logging helo}.

    \item [queueid] The queueid of the mail if the connection represents an
        accepted mail, or \texttt{NOQUEUE} otherwise.

    \item [start] The timestamp of the first log line, in seconds since the
        epoch.\glsadd{Epoch}

    \item [end] The timestamp of the last log line, in seconds since the
        epoch.

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an \acronym{SMTP} command
        --- a mechanism that is quite useful for testing new restrictions.
        This field will be set to 1 if the log line parsed was a warning rather
        than a real rejection, or to 0 for a real rejection.  This field
        should be ignored if the result is not a rejection, i.e.\ the
        action field of the associated rule is not
        \texttt{DELIVERY\_REJECTED}.

    \item [smtp\_code] The \acronym{SMTP} code associated with the log
        line.  In general an \acronym{SMTP} code is only present in a log
        line for a rejection or final delivery; results whose log line did
        not contain an \acronym{SMTP} code will duplicate the
        \acronym{SMTP} code of other results in that connection.  Some
        final delivery log lines do not contain an \acronym{SMTP} code: in
        those cases the \acronym{SMTP} code is specified by the rule's
        \texttt{result\_data} field, based on the success or failure
        represented by the log line.

    \item [enhanced\_status\_code] The enhanced status code~\cite{RFC3463}
        is similar to the \acronym{SMTP} code, but is intended to be parsed
        by mail clients so that error messages can be clearly conveyed to
        the user.  Enhanced status code support was added to Postfix in
        version 2.3; logs generated by previous versions will not contain
        any enhanced status codes.

    \item [sender] The sender's email address.  Because multiple mails may
        be delivered during one connection, there may be different sender
        addresses in the results for one connection; however there should
        not be different sender addresses in the results for one mail.
        Mails sent over the same connection can be distinguished by their
        queueids.

@@

    \item [size] The size of the mail; it will only be present for
        delivered mails.

@@

    \item [message\_id] The message-id of the accepted mail, or
        \texttt{NULL} if no mail was accepted.

    \item [data] A field available for anything not covered by other
        fields, e.g.\ the rejection message from a \acronym{DNSBL}\@.

    \item [timestamp] The time at which the log line was logged, in seconds
        since the epoch.\glsadd{Epoch}

@@

Each time the parser is run the framework performs some initialisation
tasks, setting up data structures that will later be used by actions,
including several state tables.  Data about the connections and mails being
processed is held in \texttt{connections} and \texttt{queueids}
respectively.  There are also state tables used when working around
complications in parsing: \texttt{timeout\_queueids} is used when dealing
with connections that time out during the DATA phase (\sectionref{timeouts
during data phase}); \texttt{bounce\_queueids} is part of the solution to
bounce notification mails being delivered before their creation is logged
(\sectionref{Bounce notification mails delivered before their creation is
logged}); and \texttt{postsuper\_deleted\_queueids} caches information
about mails recently deleted by the administrator so that subsequent log
lines processed by the SAVE\_DATA action can be discarded (\sectionref{Mail
deleted before delivery is attempted}).

@@

    \item Ensuring the regex is valid and compiling it for efficiency
        (\sectionref{Caching compiled regexes}).

@@

If there are state tables from a previous \parsername{} run to be loaded,
they will be loaded now; obviously the framework supports saving its state
tables too.  The need to track re-injected mail (\sectionref{Re-injected
mails}) complicates the process of saving state, as the relationships
between mails must be maintained; loading state is complicated by dealing
with aborted delivery attempts (\sectionref{aborted delivery attempts}),
because a different set of relationships between connections and mails must
be re-created when the state tables are restored.

The framework will now move on to parsing each log file.  For each log line
it will use those rules whose \texttt{program} field matches the program in
the log line (see \sectionref{rule conditions in implementation}), falling
back to generic rules if necessary, and warning if the log line is
unrecognised.  The rules are sorted for increased efficiency; see
\sectionref{rule ordering for efficiency} for details.  The repetitive
nature of log files gives them high compression ratios; the framework
supports reading compressed files, simplifying the process of parsing them
because they do not have to be uncompressed to a temporary file before
parsing begins.  The framework also displays a progress bar to show how far
parsing has progressed through the log file, and to give the user an
indication of how long the remainder of the parsing process is likely to
take.  The progress bar is not as reliable when parsing compressed files
because a compressed block may uncompress to a variable number of log
lines; there is also variation between the recognition and processing time
of individual log lines, but overall the progress bar is a useful addition.

The framework collects data used when evaluating \parsernames{} efficiency
for the results chapter (\sectionref{Results}); in addition some of the
techniques used to improve parsing speed can be turned off or altered to
show the effect they have.  Three sets of data are collected:

@@

    \item The framework can record which rule recognised each log line, and
        then on a subsequent run use that information so that it uses the
        correct rule for each log line.  This gives the best possible
        running time, as only one rule is used for each log line.  A
        variation of this is to use the correct rule last, to get the worst
        possible running time.  \sectionref{perfect rule ordering}
        discusses this in detail.

    \item Each regex is compiled and the result cached when the rules are
        loaded; this can be disabled and the regex compiled every time it
        is used when recognising log lines, to demonstrate the effect this
        optimisation has on parser efficiency, as shown in
        \sectionref{Caching compiled regexes}.

    \item Normally once a log line has been recognised, the framework
        invokes the action specified by the rule.  Invoking the action can
        be skipped if desired, so the timing information shows how long is
        spent recognising log lines; this time can be subtracted from the
        time taken by a normal run to show how long is spent processing log
        lines.  This data is analysed in \sectionref{recognising vs
        processing}.

    \item The framework can skip inserting data into the database, because
        that dominates the execution time of the parser, and the evaluation
        in \sectionref{Results} is measuring the speed of \parsername{},
        not the speed of the database or the disks it resides on.

@@

    \item Each result can be extended with extra debugging information, so
        that when a mail's data is dumped for inspection more information
        is available.  The information currently added is: the log line's
        timestamp in human readable form; the entire log line; the name of
        the log file and the number of the log line within it.

@@

This section explains the actions found in \parsername{}; it may help to
revisit the flow chart in \sectionref{flow chart} to see how a mail passes
from one action to another as its log lines are recognised.  The words
\textit{mail\/} and \textit{connection\/} are used in the actions
descriptions below because they are less unwieldy and more specific than
\textit{state table entry\/}; a connection becomes a mail during the
\texttt{CLONE} action, which processes Postfix accepting a delivery
attempt, and the data structure moves from one state table to another
within the parser.

The complications and difficulties that arose when parsing log files
generated on a production mail server are documented in
\sectionref{complications}; some action descriptions refer to specific
difficulties they address.  The complications are documented in a separate
section to avoid overwhelming the action descriptions.

@@

            \item [program] The name of the program that logged the line.

            \item [pid] The \acronym{pid} of the program that logged the
                line.

@@

    \item [BOUNCE\_CREATED] Postfix 2.3 and subsequent versions log the
        creation of bounce messages.  This action creates a new mail if
        necessary; if the mail already exists the unknown origin flag will
        be removed.  The action also marks the mail as a bounce
        notification.  To deal with complication \sectionref{Bounce
        notification mails delivered before their creation is logged} this
        action checks a cache of recent bounce mails, to avoid creating
        bogus bounce mails when log lines are out of order.

@@

    \item [DELETE] Deals with mail deleted using Postfix's administrative
        command \daemon{postsuper}.  This action adds a dummy recipient
        address if required, then invokes the COMMIT action to save the
        mail to the database.  The complication this action deals with is
        described fully in \sectionref{Mail deleted before delivery is
        attempted}.

>>>>

XXX make log-files-lines

XXX grep matches logparser*.tex

@@

XXX DID I SAY THE STUFF IN THE NEXT TWO PARAGRAPHS ALREADY\@?

@@

    \item Postfix logs enough information to make it possible to accurately
        reconstruct the actions it has taken.  Heuristics are used in
        several places when parsing; see \sectionref{identifying bounce
        notifications}, \sectionref{aborted delivery attempts}, and
        \sectionref{pickup logging after cleanup} for details.  At least
        one difficulty encountered while writing \parsername{}
        (\sectionref{yet more aborted delivery attempts}) could not be
        solved using the data in the log files, and requires a brute force
        approach.

@@

In some ways this task is similar to reverse engineering or replicating a
black box system based solely on its inputs and outputs.  Thus far
analysis of the log files has been sufficient to reconstruct Postfix's
behaviour, but for more difficult programs the techniques described in
\cite{black-box-error-reporting} may be useful.  There are some advantages
to treating Postfix as a black box during parser development:

@@

    \item \parsername{} is developed using real world log files rather than
        the idealised log files someone would naturally envisage when
        reading the source code, which cannot accurately communicate the
        variety of orderings in which log lines are found in log files.

    \item The parser acts as a second source of information about Postfix's
        operation, based on empirical evidence.  A separate project could
        compare the empirical knowledge inherent in \parsername{} with
        Postfix's documentation and source code to see how closely the two
        agree.

@@

A picture is said to be worth a thousand words: the flow chart in
\figureref{flow chart image} shows the most common paths a connection or
mail can take through \parsername{}; the difficulties described in
\sectionref{complications} are excluded for the sake of clarity.  The flow
chart is intended to be a graphical overview of how a mail progresses
through both Postfix and \parsername{}, providing an overall context into
which the detailed descriptions in the remainder of this chapter will fit,
in particular the actions (\sectionref{actions in implementation}) and
complications (\sectionref{complications}).

@@

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop}, by \acronym{SMTP}, by re-injection due to
forwarding, or internally generated by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately \flowchart{PICKUP}{2}, and
the mail moves on to the delivery stage.  Re-injection due to forwarding
sadly lacks explicit log lines of its own; it is explained fully in
\sectionref{Re-injected mails}.  Internally generated mails lack any
explicit origin in Postfix 2.2.x and must be detected using heuristics as
described in \sectionref{identifying bounce notifications}; later versions
of Postfix do provide log lines for internally generated mails.  Bounce
notifications are the primary example of internally generated mails, though
there are other types, e.g.\ Postfix may generate mails to the
administrator when it encounters configuration errors, but such mails are
presumably rare.  XXX WOULD THIS PARAGRAPH BE BETTER AS A LIST\@?

@@

    \item The client disconnects \flowchart{DISCONNECT}{6}.  If Postfix has
        rejected any \acronym{SMTP} commands the resulting data will be
        saved to the database; if there were no rejections there will not
        be any data to save.  Any mails accepted will already have been
        cloned so their data is in another data structure, nd will be
        delivered in the same way as mails that entered the system by any
        other route.

@@

    \item Postfix will log the sender and recipient addresses separately
        \flowchart{SAVE\_DATA}{8}.

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail \flowchart{TRACK}{9}; the handling of
        re-injected mails is described in \sectionref{tracking re-injected
        mail}.

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator \flowchart{COMMIT}{11}.  This is the last log line
        for this particular mail, though it may be indirectly referred to
        if it was re-injected.  The mail is cleaned up and entered in the
        database, then deleted from the state tables.

@@

A database is used to store both the rules and the data gleaned by parsing
Postfix log files.  Understanding the database schema is important in
understanding the actions of the parser, and essential to developing
further applications which utilise the data gathered; \sectionref{database
as API} describes how the database schema functions as an \acronym{API}.

@@

The data saved from parsing the log files is also divided into two tables:
connections and results.  The connections table contains an entry for every
mail accepted and every connection where there was a rejection; the
individual fields will be described in \sectionref{connections table}.  The
results table will have at least one entry for each entry in the
connections table; its fields will be covered in detail in
\sectionref{results table}.  A diagram of the database schema is provided
in \figureref{Diagram of the Database Schema picture}, to complement the
in-depth descriptions of each table.

@@

The database populated by \parsername{} provides a simple interface to
Postfix log files.  Although the interface is a database schema rather than
a set of functions in a library, it gives the same benefits as any other
\acronym{API}: it provides a stable interface between the user and the
creator of the data, allowing code on either side of the interface to be
changed without adverse effects, as long as the interface is adhered to.
Programs that use the database can range from the simple examples in
\sectionref{motivation} to far more complex data mining tools.

@@

\begin{verbatim}
CREATE TRIGGER ON INSERT INTO results
    WHERE sender = "boss@example.com"
        AND action = "DELIVERY_REJECTED"
    SEND PANIC EMAIL TO "postmaster@example.com";
\end{verbatim}

            \item Other tables can be added to the database, e.g.\ to cache
                historical or computed data, or to link in data from other
                sources.

@@

    \item \acronym{SQL} is reasonably standard and many people will already
        be familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn, e.g.\
        \urlLastChecked{http://philip.greenspun.com/sql/}{2009/02/23}.
        Although every vendor implements a different dialect of
        \acronym{SQL}, the basics are the same everywhere, analogous to the
        overall similarity and minor differences amongst Irish English,
        British English, American English, and Australian English.
        Depending on the database in use there may be tools available that
        reduce or remove the requirement to know \acronym{SQL}.

@@

Rules are discussed in detail in \sectionref{rules in implementation}, but
the structure of the rules table is documented here along with the other
tables in the database.  Rules are created by the user, not the parser, and
will not be modified by \parsername{} (except for the hits and hits\_total
fields).  Rules classify the individual log lines, capturing data to be
saved in the connections and results tables, and specifying the action to
take for each recognised log line.

Each rule some or all of these fields; all fields are required unless
otherwise stated.

@@

    \item [description] Something must have occurred to cause Postfix to
        log each line, this field should describe the event causing the log
        lines this rule recognises.

    \item [restriction\_name] The name of the Postfix restriction that
        caused the mail to be rejected.  Only applicable to rules that
        recognise rejection log lines, i.e.\ rules that have an action of
        \texttt{DELIVERY\_REJECTED}; other rules should not set this field.

@@

    \item [regex] The regex to recognise log lines with, as documented in
        \sectionref{regex components}.

    \item [connection\_data] Sometimes rules need to provide data that is
        not present in the log line: e.g.\ setting \texttt{client\_ip} when
        a mail is being delivered to another server.  Any field in the
        connections table can be set in this way.  The format is:
        \newline{} \tab{} \texttt{ client\_hostname = localhost,}
        \newline{} \tab{} \tab{} \texttt{client\_ip = 127.0.0.1} \newline{}
        i.e.\ semi-colon or comma separated assignment statements.  Commas
        and semi-colons cannot be escaped and thus cannot be included in
        data, this feature is intended for use with small amounts of data,
        so dealing with escape sequences was deemed unnecessary.  This
        field is optional.

    \item [result\_data] The result table equivalent of
        \texttt{connection\_data}, also optional.

@@

    \item [hits] This counter is maintained for every rule and incremented
        each time the rule successfully recognises a log line.  At the
        start of each run \parsername{} sorts the rules by hits, and at the end of the run it updates
        every rule's hits field in the database.  Assuming that the
        distribution of log lines is reasonably consistent across log
        files, ordering rules by their recognition frequency will reduce
        the parser's execution time.  Rule ordering for efficiency is
        discussed in \sectionref{rule ordering for efficiency}.  This field
        will be set by the parser rather than the rule author.

    \item [hits\_total] The total number of hits for this rule over all
        runs of the parser; hits is reset to zero each time the parser is
        run, but hits\_total is not.  This field will be set by the parser
        rather than the rule author.

    \item [priority] This is the user-configurable companion to hits: when
        the list of rules is sorted, priority overrides hits.  This allows
        more specific rules to take precedence over more general rules, as
        described in \sectionref{rules in architecture}.  This field is
        optional.

@@

Every accepted mail and every connection where there was a rejection will
have a single entry in the connections table containing all of the
following fields:

@@

    \item [server\_ip] The \acronym{IP} address of the server: the local
        address when receiving mail, the remote address when sending mail.

@@

    \item [client\_ip] The client \acronym{IP} address: the remote address
        when receiving mail, the local address when sending mail.

@@

    \item [helo] The hostname used in the HELO command.  The HELO hostname
        occasionally changes during a connection, presumably because spam
        or virus senders think it is a good idea.  By default Postfix only
        logs the HELO hostname when it rejects an \acronym{SMTP} command,
        but it is quite simple to rectify this as described in
        \sectionref{logging helo}.

    \item [queueid] The queueid of the mail if the connection represents an
        accepted mail, or \texttt{NOQUEUE} if not.

    \item [start] The timestamp of the first log line.

    \item [end] The timestamp of the last log line.

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an \acronym{SMTP} command
        --- a mechanism that is quite useful for testing new restrictions.
        This field will be false for a real rejection, or true if the log
        line was a warning.  This field should be ignored if the result is
        not a rejection, i.e.\ the action field of the associated rule is
        not \texttt{DELIVERY\_REJECTED}.

    \item [smtp\_code] The \acronym{SMTP} code associated with the log
        line.  In general an \acronym{SMTP} code is only present in a log
        line representing a rejection or final delivery; results whose log
        line did not contain an \acronym{SMTP} code will duplicate the
        \acronym{SMTP} code of other results in that connection.  Some
        final delivery log lines do not contain an \acronym{SMTP} code: in
        those cases the \acronym{SMTP} code is specified by the rule's
        \texttt{result\_data} field, based on the success or failure
        represented by the log line.  XXX GIVE AN EXAMPLE LOG LINE\@.

    \item [enhanced\_status\_code] The enhanced status code~\cite{RFC3463}
        is similar to the \acronym{SMTP} code, but is intended to be
        interpreted by mail clients so that error messages can be clearly
        conveyed to the user.  Enhanced status code support was added to
        Postfix in version 2.3; log files from previous versions will not
        contain any enhanced status codes.

    \item [sender] The sender's email address.  This may change during a
        connection if the client uses different sender addresses for
        multiple rejected delivery attempts; however there should not be
        different sender addresses in the results for one accepted mail.
        Mails sent over the same connection can be distinguished by their
        queueids.

@@

    \item [size] The size of the mail, present for delivered mails only.

@@

    \item [message\_id] The message-id of the accepted mail, again present
        for delivered mails only.

    \item [data] A field available to store a piece of captured data that
        does not have a specific field, e.g.\ the rejection message from a
        \acronym{DNSBL}\@.

    \item [timestamp] The time at which the log line was logged.

@@

Each time the parser is run the framework performs some initialisation
tasks, setting up data structures that will later be used by actions,
including several state tables.  Data about the connections and mails being
processed is held in \texttt{connections} and \texttt{queueids}
respectively.  The other state tables used when working around
complications in parsing: \texttt{timeout\_queueids} is used when dealing
with connections that time out during the DATA phase (\sectionref{timeouts
during data phase}); \texttt{bounce\_queueids} is part of the solution to
bounce notification mails being delivered before their creation is logged
(\sectionref{Bounce notification mails delivered before their creation is
logged}); and \texttt{postsuper\_deleted\_queueids} caches information
about mails recently deleted by the administrator so that subsequent log
lines processed by the SAVE\_DATA action can be discarded (\sectionref{Mail
deleted before delivery is attempted}).

@@

    \item The regex is valid and compiling it for efficiency
        (\sectionref{Caching compiled regexes}).

@@

If there are state tables from a previous \parsername{} run to be loaded,
they will be loaded now; the framework supports saving state at any time,
without adversely affecting the parsing process.  The need to track
re-injected mail (\sectionref{Re-injected mails}) complicates the process
of saving state, as the relationships between mails must be maintained;
loading state is also complicated by dealing with aborted delivery attempts
(\sectionref{aborted delivery attempts}), because a separate set of
relationships between connections and mails must be re-created when the
state tables are restored.

The framework is now ready to begin parsing.  For each log line it will use
those rules whose \texttt{program} field matches the program in the log
line (see \sectionref{rule conditions in implementation}), falling back to
generic rules if necessary, and warning if the log line is unrecognised.
The rules are sorted for increased efficiency; see \sectionref{rule
ordering for efficiency} for details.  The repetitive nature of log files
gives them high compression ratios; the framework supports reading
compressed files, simplifying the process of parsing them because they do
not have to be uncompressed to a temporary file before parsing begins.  The
framework also displays a progress bar to show how far parsing has
progressed through the log file, and to give the user an indication of how
long the remainder of the parsing process is likely to take.  The progress
bar is not as reliable when parsing compressed files, because a compressed
block may uncompress to a variable number of log lines; there is also
variation between the recognition and processing time of individual log
lines, but overall the progress bar is a useful addition.

The framework collects data used when evaluating \parsernames{} efficiency
for the results chapter (\sectionref{Results}); some of the techniques used
to improve parsing speed can be turned off or altered to measure the effect
they have.  Three sets of data are collected:

@@

    \item The framework can record which rule recognised each log line, and
        then on a subsequent run consult that information so that it uses
        the correct rule for each log line.  This gives the best possible
        running time, because only one rule is used for each log line.  A
        variation of this is to use the correct rule last, to get the worst
        possible running time.  \sectionref{perfect rule ordering}
        discusses this in detail.

    \item Each regex is compiled and the result cached when the rules are
        loaded; this can be disabled and the regex compiled every time it
        is used when recognising log lines.  The effect of this
        optimisation is shown in \sectionref{Caching compiled regexes}.

    \item Normally when a log line has been recognised the framework
        invokes the action specified by the rule.  Invoking the action can
        be skipped if desired, so the timing information shows how long is
        spent recognising log lines only; this time can be subtracted from
        the time taken by a normal run to show how long is spent processing
        log lines.  This data is analysed in \sectionref{recognising vs
        processing}.

    \item The framework can skip inserting data into the database, because
        that dominates the execution time of the parser; the evaluation
        chapter measures the speed of \parsername{}, not the speed of the
        database or the disks it resides on.

@@

    \item Each result can be extended with extra debugging information,
        which is useful when a warning dumps a mail's data for inspection.
        The information currently added is: the log line's timestamp in
        human readable form; the entire log line; the name of the log file
        and the number of the log line within it.  This information is not
        stored in the database.

@@

This section explains the actions found in \parsername{}; it may help to
revisit the flow chart in \sectionref{flow chart} to see how a mail passes
from one action to another as its log lines are recognised.  The words
\textit{mail\/} and \textit{connection\/} are used in the actions
descriptions below because they are less unwieldy and more specific than
\textit{state table entry\/}; a connection becomes a mail during the
\texttt{CLONE} action, which processes Postfix accepting a delivery
attempt, and the data structure moves from the connections state table to
the queueids state table.

The complications and difficulties that arose when parsing real-world log
files are documented in \sectionref{complications}; some action
descriptions refer to specific difficulties they address.  The
complications are documented in a separate section to avoid overwhelming
the action descriptions.

@@

            \item [program] The name of the program that generated the log
                line.

            \item [pid] The \acronym{pid} of the process that logged the
                line.

@@

    \item [BOUNCE\_CREATED] Postfix 2.3 and subsequent versions log the
        creation of bounce messages.  This action creates a new mail if
        necessary; if the mail already exists the unknown origin flag will
        be removed from it.  The action also marks the mail as a bounce
        notification.  To deal with complication \sectionref{Bounce
        notification mails delivered before their creation is logged} this
        action checks a cache of recent bounce mails, to avoid creating
        bogus bounce mails when log lines are out of order.

@@

    \item [DELETE] Deals with mail deleted using Postfix's administrative
        command \daemon{postsuper}.  This action adds a dummy recipient
        address if required (see \sectionref{Mail deleted before delivery
        is attempted}), then invokes the COMMIT action to save the mail to
        the database.
